\renewcommand{\chapcolor}{chGitlab}
\section{GitLab Pipelines}
\subsection{Core Concepts \& DSL Mapping}
Pipeline = automated steps $\rightarrow$ faster, safer, reliable. Traceable logs, parallel jobs, consistent envs kill "works on my machine".
\begin{center}\footnotesize
\begin{tabular}{@{}lll@{}}
\textbf{Concept} & \textbf{Keyword} & \textbf{Example} \\
\hline
Stages & \texttt{stages:} & \texttt{[build, test, deploy]} \\
Jobs & top-level key & \texttt{<name>: \{stage:, script:\}} \\
Steps & \texttt{script:} & \texttt{["cmd1","cmd2"]} \\
Rules & \texttt{rules}/\texttt{only}/\texttt{except} & \texttt{[\{if:..., when:...\}]} \\
Services & \texttt{services:} & \texttt{[\{name: postgres:15, alias: db\}]} \\
Runners & \texttt{tags:} (UI config) & \texttt{["my-runner"]} \\
Caches & \texttt{cache:} & \texttt{\{key, paths\}} (deps, speedup) \\
Artifacts & \texttt{artifacts:} & \texttt{\{paths, expire\_in\}} (persist) \\
Registries & \texttt{script:} push & \texttt{docker login/build/push} \\
Variables & \texttt{variables:} & \texttt{\{FOO: bar\}} (mask/protect UI) \\
\end{tabular}\end{center}
\textbf{5-stage flow:} commit/MR/tag $\rightarrow$ \textit{build} (compile, image, package) $\rightarrow$ \textit{test} (unit/integration/security/perf) $\rightarrow$ \textit{release} (publish to registry) $\rightarrow$ \textit{deploy} (target env) $\rightarrow$ \textit{post-deploy} (smoke, health, rollback).

\subsection{Jobs, Stages \& Rules}
\texttt{.gitlab-ci.yml} YAML DSL. \textbf{Job} = smallest unit, belongs to one \textbf{stage}, runs \texttt{script} on a runner. Same stage $\rightarrow$ \textit{parallel}; stages $\rightarrow$ \textit{sequential}. \\
\textbf{rules:} \textit{when} a job runs (modern \texttt{only/except}). \textbf{needs:} DAG, start before stage finishes. \textbf{dependencies:} which jobs' artifacts to pull (no ordering). \textbf{workflow:rules:} gate the whole pipeline.
\begin{lstlisting}[language=yaml]
workflow:
  rules:
    - if: $CI_PIPELINE_SOURCE == "merge_request_event"
    - if: $CI_COMMIT_BRANCH && $CI_OPEN_MERGE_REQUESTS
      when: never
    - if: $CI_COMMIT_BRANCH
deploy:
  stage: deploy
  needs: [build]           # DAG, skip stage wait
  dependencies: [build]    # pull build's artifacts
  rules:
    - if: '$CI_COMMIT_BRANCH == "main"'
    - when: never
\end{lstlisting}

\subsection{Cache, Artifacts \& Services}
\textbf{Cache:} shared across jobs/pipelines, speed up repeat work (deps, compiled objects). \textbf{Artifacts:} files persisted to GitLab for downstream stages or download; declare via \texttt{dependencies}/\texttt{needs}. \textbf{Services:} sidecar containers (DB, dind, API) reachable via \texttt{alias}.
\begin{lstlisting}[language=yaml]
integration-test:
  services: [{name: postgres:17, alias: db}]
  variables: {POSTGRES_DB: app, POSTGRES_USER: u, POSTGRES_PASSWORD: p}
  script: [psql -h db -U $POSTGRES_USER -d $POSTGRES_DB -c "SELECT 1"]
  cache: {key: deps, paths: [.cache/]}
  artifacts: {paths: [bin/], expire_in: 1h}
\end{lstlisting}

\subsection{Environments}
Where code is deployed. \textbf{Tiers:} Local (workstation) $\rightarrow$ Dev (sandbox) $\rightarrow$ Integration (CI target) $\rightarrow$ Test/QC (interface tests) $\rightarrow$ Staging (prod mirror) $\rightarrow$ Production (live). Linkable to K8s cluster (set up via UI). \texttt{environment.on\_stop} unstages; \texttt{action: stop} stops it.

\subsection{Push vs. Pull Deployment}
\textbf{Push} (Jenkins/GitLab calls cluster): + easy, flexible targets (VMs/containers, not just K8s); -- firewall open, admin access out, pipeline per env. \textbf{Pull} (agent in cluster, e.g. Argo/Flux): + secure (no inbound), auto-reconcile, scales identically; -- agent per cluster, polling latency.

\subsection{GitLab Runner}
Worker executing jobs. \textbf{Scope:} Instance (shared, GitLab-managed) / Group / Project (dedicated). \textbf{Executors:} \texttt{shell}, \texttt{ssh}, \texttt{docker}, \texttt{docker+machine}, \texttt{kubernetes}, \texttt{virtualbox}, \texttt{parallels}. Register: \texttt{gitlab-runner register} with URL + token + tags + executor. Jobs pick runner by \texttt{tags:}.

\subsection{More Keywords \& Docker Build}
\texttt{default:} fallback \texttt{image}/\texttt{before\_script}. \texttt{before\_script}/\texttt{after\_script} wrap every \texttt{script}. \texttt{rules:changes:} run only if files changed. \\
\textbf{Predefined vars:} \texttt{\$CI\_REGISTRY[\_IMAGE]}, \texttt{\$CI\_REGISTRY\_USER}/\texttt{\_PASSWORD}, \texttt{\$CI\_COMMIT\_BRANCH}, \texttt{\$CI\_COMMIT\_SHORT\_SHA}, \texttt{\$CI\_PROJECT\_DIR}, \texttt{\$CI\_JOB\_TOKEN} (auth API/registry), \texttt{\$CI\_MERGE\_REQUEST\_IID}. Pin images by digest (\texttt{img@sha256:..}), not \texttt{:latest} $\rightarrow$ flaky.
\begin{lstlisting}[language=yaml]
build-image:
  stage: release
  image: docker:28
  services:
    - {name: docker:28-dind, alias: docker}   # Docker-in-Docker
  variables: {IMG: $CI_REGISTRY_IMAGE:$CI_COMMIT_SHORT_SHA}
  before_script:
    - docker login -u $CI_REGISTRY_USER -p $CI_REGISTRY_PASSWORD $CI_REGISTRY
  script:
    - docker build -t $IMG -f Dockerfile.prod .
    - docker push $IMG
  rules:
    - changes: [Dockerfile.prod]
\end{lstlisting}
