\renewcommand{\chapcolor}{chHelm}
\section{Helm}
Package manager for K8 (apt/brew-like). \textbf{Chart}=files describing related K8 resources. \textbf{Release}=running instance. \textbf{Repo}=stores versioned charts. \textbf{Arch:} client-side only, talks K8 API direct, release state in K8 \textbf{Secrets} (no Tiller). \textbf{Pros:} huge chart library, versioning/deps, lifecycle (install/upgrade/rollback), powerful Go-template logic. \textbf{Cons:} templates hard to debug, "spaghetti YAML", opaque state, steep curve. Use for: distributing complex apps, consuming community charts, advanced logic. Avoid for: simple YAML tweaks.

\subsection{Commands}
\begin{lstlisting}[language=sh]
helm repo add|list|rm <repo> <url>          # repos
helm search repo <r>/<c> -l                 # all versions
helm install -f vals.yaml <rel> <chart> --version 1.2.3
helm upgrade <rel> <chart>; helm rollback <rel> <rev>
helm uninstall <rel>           # also drops history
helm list; helm history <rel>  # CHART vs APP VERSION
helm status <rel>              # unknown|deployed|uninstalled|
  # superseded|failed|uninstalling|pending-{install,upgrade,rollback}
helm get all|values|notes|manifest <rel>    # values=overrides only
helm template <n> <c> -f v.yaml --set k=v   # render local
helm install <n> <c> --dry-run --debug      # simulate
helm lint <chart> [--with-subcharts]
helm show chart|readme|values|all <chart>
helm package <dir>             # build .tgz; helm push <pkg> <repo>
helm dependency update         # pulls into charts/
\end{lstlisting}

\subsection{Chart Structure \& Chart.yaml}
\texttt{helm create <n>} scaffolds: \texttt{Chart.yaml} (meta), \texttt{values.yaml} (defaults), \texttt{templates/} (Go-tmpl manifests), \texttt{templates/\_helpers.tpl} (named templates, \texttt{\_}=not rendered), \texttt{templates/NOTES.txt} (rendered post-install), \texttt{charts/} (subcharts), \texttt{crds/}, \texttt{.helmignore} (globs, \texttt{!}=negate; excludes from \texttt{helm package}).\\
\texttt{Chart.yaml} fields: \textbf{apiVersion} (\texttt{v2}=Helm 3+, \texttt{v1}=legacy), \textbf{name}, \textbf{version} (SemVer2; pkg name = name+ver), \textbf{type} (\texttt{application} default \textbar{} \texttt{library} = helpers only, no install), \textbf{appVersion} (app being shipped, info-only), \textbf{kubeVersion} (range like \texttt{>=1.12.1 <1.14.0 || >=1.14.2}; install fails if mismatch), \textbf{dependencies} (name/version/repository/condition/alias; \texttt{requirements.yaml} deprecated, lockfile on update).

\subsection{Templating}
\textbf{Top-level objects:} \texttt{.Release.\{Name,Namespace,Revision,IsInstall,IsUpgrade,Service\}}, \texttt{.Chart.*}, \texttt{.Values}, \texttt{.Files.Get <path>}, \texttt{.Capabilities.\{APIVersions,KubeVersion\}}, \texttt{.Template.\{Name,BasePath\}}. \\
\textbf{Values precedence} (low$\to$high, last wins): subchart \texttt{values.yaml} $\to$ parent \texttt{values.yaml} $\to$ \texttt{-f myvals.yml} $\to$ \texttt{--set k=v}. Unset default: \texttt{--set x=null}. \\
\textbf{Functions:} Sprig lib (\textasciitilde 60+: \texttt{upper quote eq default trunc toYaml indent nindent}). Pipelines \texttt{|} pass result as last arg. \texttt{include} = \texttt{template} but pipeable. \texttt{required "msg" .Values.x} fails if empty. \texttt{tpl <str> .} evaluates string as template. \\
\textbf{Flow:} \texttt{\{\{ if/else if/else \}\}...\{\{ end \}\}} (false on 0/""/nil/empty); \texttt{\{\{ range \$k,\$v := .Values.list \}\}}; \texttt{\{\{ with .Values.sub \}\}} rescopes \texttt{.}. Vars: \texttt{\{\{- \$n := .Release.Name -\}\}}. \\
\textbf{Whitespace:} \texttt{\{\{-} strips left, \texttt{-\}\}} strips right; \texttt{nindent N} = newline + indent. YAML indent critical.
\begin{lstlisting}[language=yaml]
{{- define "app.name" -}}{{ default .Chart.Name .Values.name | trunc 63 }}{{- end }}
name: {{ include "app.name" . | quote }}
{{- required "set .Values.img!" .Values.img }}
resources: {{- toYaml .Values.resources | nindent 4 }}
\end{lstlisting}
