\renewcommand{\chapcolor}{chMonitor}
\section{K8s Monitoring \& Logging}
\textbf{Observability} = property (degree system generates actionable insights, qualitative); \textbf{Observing}/Monitoring = process (collect/store/evaluate \textit{signals}). \textbf{Three pillars:} \textit{Metrics} (quantitative: util, counters, rates), \textit{Logs} (qualitative: stack traces, events), \textit{Traces} (request path across services via \textit{spans} -- latency/bottlenecks, e.g.\ Tempo, Jaeger). \textbf{Monitoring vs Logging} complement each other. \\
\textbf{K8s obs gaps:} no historical metrics, \texttt{kubectl logs} only running containers (lost when container dies, kubelet cleanup), no native dashboards/alerts. Container logs at \texttt{/var/log/pods/...}; kubelet rotates at 10MB, keeps 5. \\
\textbf{Push (agent-based):} agent on system sends to collector; observed decides when. \textbf{Pull (agentless):} collector queries via SNMP/ICMP/HTTP; collector decides, but system must be online. \textbf{Golden Signals} (Google SRE): latency, traffic, errors, saturation.
\begin{mdframed}
\textbf{PLG stack} (lightweight, K8s-native): Prometheus+Loki+Grafana. \textbf{LGTM:} Loki, Grafana, Tempo, Mimir. \textbf{EFK/ELK:} Elasticsearch (Java, JSON index, full-text, heavy) + Kibana + Fluentd/Logstash. PLG indexes \textit{labels} only (cheap, fast, weaker text search). \textbf{OpenTelemetry:} unified telemetry (metrics/logs/traces) spec.
\end{mdframed}
\subsection{Prometheus \& PromQL}
Go, TSDB, pull-based scrape over HTTP (\texttt{/metrics}). Pushgateway/remote-write for short jobs. Client libs Go/Java/Python/Ruby/Rust. Standalone (no remote dep). \\
\textbf{Targets:} Kubelets, kube-state-metrics, exporters. \textbf{Exporters:} bridge for apps without native \texttt{/metrics} (PostgreSQL, NGINX, Redis, JIRA, Kafka, \dots); use when can't instrument. \textbf{node-exporter:} HW/OS. \textbf{kube-state-metrics:} K8s object state. Install \texttt{metrics-server} for \texttt{kubectl top}. CPU units: \texttt{100m}=0.1 CPU absolute.\\
\textbf{Metric types:} \textit{Counter} (monotonic, reset to 0: requests/errors), \textit{Gauge} (current value: memory, concurrency), \textit{Histogram} (sampled buckets), \textit{Summary} (sampled + total/sum). \textbf{Format:} \texttt{name\{label=val,\dots\} value [ts]} -- foundation of \textbf{OpenMetrics} std. \\
\textbf{PromQL data types:} \textit{Instant vector} (set of series, one sample each, same ts), \textit{Range vector} (range per series, e.g.\ \texttt{[5m]}), \textit{Scalar} (float), \textit{String} (unused). \textbf{Offset} (positive = backwards): \texttt{rate(x[5m] offset 1w)}. \textbf{@-modifier}: \texttt{sum(x \@ 1609746000)}. \texttt{histogram\_quantile(0.95, rate(\dots\_bucket[5m]))} for p95 from Histogram. \texttt{rate()} useless on spikes/outliers -- use raw \& \texttt{>} threshold.
\begin{lstlisting}[language=sql]
sum by (namespace) (kube_pod_status_ready{condition="false"})
rate(mysql_global_status_commands_total{command=~"select"}[5m])*100 > 35
\end{lstlisting}
\textbf{Prometheus Operator} (\texttt{kube-prometheus-stack}): bundles operator+Prom+Grafana+Alertmanager. CRDs \texttt{Prometheus}/\texttt{ServiceMonitor}/\texttt{PodMonitor}. Match chain: Prom \texttt{serviceMonitorSelector}$\rightarrow$SM \texttt{selector}$\rightarrow$Service \texttt{selector}$\rightarrow$Pods. \\
\textbf{Scaling Prometheus:} \textit{Federation} (hierarchical: region scrapes DC); \textit{Thanos} (sidecar, global query, long-term, dedup); \textit{Grafana Mimir} (horiz.\ scalable, ingester/querier components, multi-tenant, remote-write only).

\subsection{Grafana, Alerting}
General-purpose viz, many data sources (Prom/Loki/ES). \textit{Explore} feature: browse data sources, build queries, debug, share. OAuth/SAML/LDAP. Dashboard JSON repo \url{grafana.com/grafana/dashboards}. Tips: important data top, status-light $>$ number, simplicity, Data Source variable. \\
\textbf{Alertmanager vs Grafana Alerts:} Alertmanager (Prom ecosystem, GitOps-friendly, separates viz/alert, Prom-only). Grafana Alerts (any data source, tied to dashboards, can chain to ext Alertmanager, less GitOps-friendly). Flow: \textit{Rule}$\rightarrow$evaluate/threshold$\rightarrow$\textit{Alert}$\rightarrow$Silence?$\rightarrow$AlertManager$\rightarrow$Channels (Email/Slack/Teams/Discord/Threema/Webhook)$\rightarrow$Contact Points. \textbf{Warning $\ne$ Alert}: cert expires in 45d, CPU 75\%, budget 200\% are warnings not pages. Avoid alert fatigue, build up slow, test.

\subsection{Logging: Loki, Alloy}
\textbf{Architectures:} \textit{Node-level} (DaemonSet agent reads node log file -- survives container death, batched); \textit{App-level} (app pushes to backend directly); \textit{Sidecar} (streaming = sidecar writes to file then node agent ships -- \texttt{kubectl logs} won't show sidecar's own; logging-agent sidecar = ships to backend directly, low-latency but loses on disconnect, adds complexity, misses crashes). Best practice: structured JSON logs. \\
\textbf{Loki:} horiz.\ scalable, indexes \textit{labels} only (not text), \textit{nanosecond} unix ts, cheap storage, PromQL-like \textbf{LogQL}: \texttt{\{job="test"\} |= "error"}; metric queries \texttt{rate(\{app="x"\} |= "176" [1m])}; filters \texttt{|=} contains, \texttt{!=}, \texttt{|\~{}} regex, \texttt{!\~{}}. \\
\textbf{Alloy} (Grafana, replaces Promtail): collects all telemetry, own HCL-like lang. Components: \texttt{loki.source.file} (tail files, classic DaemonSet pattern), \texttt{loki.source.kubernetes} (subscribe via K8s API -- no root, no DaemonSet needed, but more CPU/net), \texttt{loki.source.kubernetes\_events} (K8s events). Logs enriched with \textbf{K8s metadata}, forwarded over HTTP. Helm-installable. \\
\textbf{Pitfalls:} too much logging $\rightarrow$ storage/complexity; no alert tuning $\rightarrow$ noise/missed; forgetting rotation/archival; relying on single signal; logging secrets.
