\renewcommand{\chapcolor}{chK8s}
\section{Kubernetes (K8)}
\begin{mdframed}
\textbf{Object:} persistent record-of-intent. \textbf{Controller:} reconciles current $\rightarrow$ desired state.
\end{mdframed}

\subsection{Architecture}
\textbf{Control plane:} \texttt{kube-apiserver} (REST front), \texttt{etcd} (KV store), \texttt{kube-scheduler} (pod $\rightarrow$ node), \texttt{kube-controller-manager} (built-in controllers), \texttt{cloud-controller-manager} (cloud glue, optional). \textbf{Node:} \texttt{kubelet} (runs pods via container runtime), \texttt{kube-proxy} (Service networking/iptables).

\subsection{Objects}
\textbf{Pod:} smallest unit, 1+ containers share net/storage. \textbf{Sidecar:} aux container alongside main (logs, proxy). \textbf{Init container:} runs to completion before main. \textbf{ReplicaSet:} keeps N pod replicas (rarely used directly). \textbf{Deployment:} manages ReplicaSets, rolling updates, rollback. \textbf{StatefulSet:} ordered pods with stable identity + per-pod PVC (DBs, brokers). \textbf{DaemonSet:} one pod per (matching) node -- logs, CNI, monitoring. \textbf{Service:} stable VIP/DNS + LB over pod set. \textbf{Ingress:} L7 HTTP(S) entrypoint. \textbf{Volume:} dir mounted into pod. \textbf{ConfigMap:} non-secret KV config. \textbf{Secret:} sensitive KV, base64-encoded, \textit{not} encrypted by default. \textbf{HPA:} autoscales replicas on metrics.
\begin{lstlisting}[language=yaml]
apiVersion: apps/v1
kind: Deployment
metadata: { name: app }
spec:
  replicas: 3
  selector: { matchLabels: { app: web } }
  strategy:
    type: RollingUpdate
    rollingUpdate: { maxUnavailable: 1, maxSurge: 1 }
  template:
    metadata: { labels: { app: web } }
    spec:
      initContainers:
      - { name: init, image: busybox, command: ['sh','-c','sleep 10'] }
      containers:
      - name: web
        image: nginx
        ports: [{ containerPort: 80 }]
      - { name: sidecar, image: busybox, command: ['sh','-c','while true; do sleep 30; done'] }
\end{lstlisting}

\subsection{Services}
Stable VIP + DNS over label-selected pods, round-robin LB. \textbf{ClusterIP} (default) internal only; \textbf{NodePort} static port on every node \texttt{<NodeIP>:<port>}; \textbf{LoadBalancer} cloud LB w/ public IP; \textbf{ExternalName} CNAME alias to external DNS (no proxy). \textbf{Ingress} routes HTTP(S) by host/path (needs controller: Nginx, Traefik). \textbf{Gateway API} (\texttt{Gateway}+\texttt{HTTPRoute}/\texttt{TLSRoute}) -- modern, multi-protocol, role-oriented Ingress replacement. \textit{Port trap:} Service \texttt{port} vs \texttt{targetPort} (container) vs \texttt{nodePort}; Ingress backend refs Service port.

\subsection{ConfigMaps \& Secrets}
Inject single var (\texttt{configMapKeyRef}/\texttt{secretKeyRef}), all keys (\texttt{envFrom}), or mount as files.
\begin{lstlisting}[language=yaml]
env:
- name: APIKEY
  valueFrom: { configMapKeyRef: { name: cfg, key: api-key } }
envFrom:
- configMapRef: { name: cfg }
volumeMounts:
- { name: v, mountPath: /etc/cfg }
volumes:
- { name: v, configMap: { name: cfg } }  # or: secret: { secretName: s }
\end{lstlisting}

\subsection{Storage: PV, PVC, StorageClass}
Container FS is ephemeral. \textbf{Volume:} dir in pod, lives w/ pod (\texttt{emptyDir}, \texttt{hostPath}, NFS, \dots). \textbf{PersistentVolume (PV):} cluster-scoped piece of external storage. \textbf{PersistentVolumeClaim (PVC):} namespaced request for storage (size, accessMode); binds to a matching PV. \textbf{StorageClass:} dynamic provisioner -- creates a PV on-demand if no static PV matches (e.g. \texttt{gp3}, \texttt{standard-rwo}).

\subsection{LimitRange \& ResourceQuota}
\textbf{LimitRange:} per-container min/max + default req/lim in NS; out-of-range Pod rejected, missing req/lim auto-injected. \textbf{ResourceQuota:} caps summed NS usage; over-quota Pod rejected.

\subsection{Namespaces}
Logical isolation in one cluster (RBAC, quotas). Namespaced (Pods, Services) isolated; cluster-wide (Nodes, PV) shared. Defaults: \texttt{default}, \texttt{kube-system}, \texttt{kube-public}, \texttt{kube-node-lease}.

\subsection{Rolling Updates}
Deployment ensures availability during change. \textbf{Recreate:} kill all $\rightarrow$ recreate (downtime). \textbf{RollingUpdate} (default): tuned by \texttt{maxUnavailable} (max pods down) + \texttt{maxSurge} (max extra pods). \texttt{kubectl rollout history/undo} for revisions.

\subsection{Scheduling}
\texttt{kube-scheduler} picks node via \textbf{Filtering} then \textbf{Scoring}. \\
\textit{Filtering} (feasibility): PodFitsHostPorts, PodFitsResources, PodMatchNodeSelector, CheckNodeDiskPressure, CheckVolumeBinding. \\
\textit{Scoring} (rank): SelectorSpreadPriority (spread same Service/STS/RS), LeastRequestedPriority (free resources), NodeAffinityPriority.
\begin{lstlisting}[language=yaml]
spec:
  nodeSelector: { disktype: ssd }
\end{lstlisting}
\textbf{Taints} on nodes repel pods; \textbf{Tolerations} on pods allow (don't force) them onto tainted nodes -- pair w/ \texttt{nodeSelector}/affinity to pin. Effects: \textit{NoSchedule, PreferNoSchedule, NoExecute} (evict). No effect on DaemonSets. \\
CLI: \texttt{kubectl taint nodes <n> gpu=true:NoSchedule}, \texttt{kubectl label nodes <n> disktype=ssd} (trailing \texttt{-} removes).
\begin{lstlisting}[language=yaml]
tolerations:
- { key: "gpu", operator: "Equal", value: "true", effect: "NoSchedule" }
\end{lstlisting}
\textbf{cordon/drain:} \texttt{kubectl cordon <n>} marks unschedulable (existing stay); \texttt{drain <n>} also evicts pods (maintenance); \texttt{uncordon <n>} reverts.

\subsection{Probes \& Resources}
kubelet probes via \texttt{httpGet}/\texttt{tcpSocket}/\texttt{exec}. \textbf{Liveness:} fail $\rightarrow$ restart. \textbf{Readiness:} fail $\rightarrow$ remove from Service endpoints. \textbf{Startup:} guards slow boots, disables others until OK. Repeated liveness fail $\rightarrow$ exp backoff $\rightarrow$ \texttt{CrashLoopBackOff}. \textbf{requests:} reserved min (scheduling). \textbf{limits:} ceiling -- CPU throttled, memory over-limit $\rightarrow$ OOMKilled.
\begin{lstlisting}[language=yaml]
resources:
  requests: { cpu: "250m", memory: "64Mi" }
  limits:   { cpu: "500m", memory: "128Mi" }
livenessProbe:
  httpGet: { path: /healthz, port: 80 }
  initialDelaySeconds: 5
  periodSeconds: 10
readinessProbe:
  httpGet: { path: /ready, port: 80 }
\end{lstlisting}

\subsection{Commands}
Apply: \texttt{kubectl \{create|apply|replace\} -f manifest.yaml} \\
Gen YAML: \texttt{kubectl create deploy web --image=nginx --dry-run=client -o yaml > d.yaml} \\
Imperative: \texttt{kubectl run}, \texttt{scale deploy x --replicas=4}, \texttt{expose} \\
Update image: \texttt{kubectl set image deploy/x web=nginx:1.27} \\
Autoscale: \texttt{kubectl autoscale deploy x --cpu-percent=50 --min=1 --max=10} \\
Inspect: \texttt{kubectl get pods -A -o wide}, \texttt{describe node <n>}, \texttt{get nodes --show-labels} \\
Exec: \texttt{kubectl exec -it nginx-xxx -- sh} (drop \texttt{-it} for one-shot) \\
Rollout: \texttt{kubectl rollout \{status|history\} deploy x}, \texttt{rollout undo deploy x --to-revision=N} \\
Node ops: \texttt{kubectl cordon|uncordon|drain <n>} \\
Default ns: \texttt{kubectl config set-context --current --namespace=lab}
