\renewcommand{\chapcolor}{chKustomize}
\section{Kustomize}
Template-free overlay tool, built into kubectl (\texttt{-k}). \texttt{kustomization.yaml} lists \textbf{resources} + declarative \textbf{transformers} (cross-cutting fields) + \textbf{patches} (partial spec edits) + \textbf{generators}. \textbf{Pros:} pure YAML, native to kubectl, easy diff/review, easy to learn. \textbf{Cons:} less powerful than Helm, no deps mgmt, hard to package/distribute, limited conditionals. Use for: env-specific configs (dev/prod), avoiding templating. Often combined with Helm.

\subsection{Commands}
\begin{lstlisting}[language=sh]
kustomize create --resources deploy.yaml,svc.yaml   # scaffold
kustomize edit add resource|set image|add label|remove resource
kustomize build <dir>            # render to stdout
kustomize build <dir> | kubectl apply -f -
kubectl apply -k <dir>           # native
kubectl kustomize <dir>          # render only (no apply)
kubectl delete -k <dir>
kustomize cfg fmt <dir>          # format yaml
kustomize localize <dir>         # bundle remote URLs/git locally
kustomize edit fix               # auto-migrate deprecated syntax
\end{lstlisting}

\subsection{kustomization.yaml}
\textbf{Deprecated $\to$ new:} \texttt{commonLabels} $\to$ \texttt{labels}, \texttt{patchesStrategicMerge}/\texttt{patchesJson6902} $\to$ \texttt{patches}, \texttt{bases:} $\to$ inside \texttt{resources:}.
\begin{lstlisting}[language=yaml]
apiVersion: kustomize.config.k8s.io/v1beta1
kind: Kustomization
resources: [../../base, storage.yaml]     # files, dirs, URLs
namespace: prod                           # transformers
namePrefix: dev-
nameSuffix: -v1
commonAnnotations: {version: 1.0.0}
labels:                                   # replaces commonLabels
- includeSelectors: true
  pairs: {app.kubernetes.io/name: hello}
images:                                   # change tag/name w/o editing Deployment
- {name: nginx, newName: bitnami/nginx, newTag: latest}
patches:                                  # strategic or JSON6902
- target: {kind: Service, name: nginx-app}
  patch: |-
    - {op: replace, path: /metadata/name, value: nginx-server}
configMapGenerator:                       # hash suffix appended -> change forces rolling restart
- name: app-config
  literals: [LOG_LEVEL=debug]
  files: [config.env]                     # or files:
secretGenerator:                          # base64-encodes; same hash mechanism
- {name: ex-secret, literals: [user=admin, pass=secret]}
components: [../../components/mysql]
\end{lstlisting}
\textbf{ConfigMap/Secret modes:} (1) \textit{Traditional}: declare as resource $\to$ change does \textbf{not} restart pods (silent drift). (2) \textit{Generator}: \texttt{configMapGenerator}/\texttt{secretGenerator} appends content hash to name $\to$ references rewritten $\to$ pods rolling-restart on change.

\subsection{Bases, Overlays, Components}
\textbf{Base:} common (\texttt{base/} with \texttt{kustomization.yaml}). \textbf{Overlay:} env-specific dir (\texttt{overlays/dev}, \texttt{overlays/prod}) that lists base in \texttt{resources:} + adds patches/transformers $\to$ produces a \textbf{variant}. \textbf{Component:} \texttt{kind: Component} (\texttt{v1alpha1}); reusable mix-in chunk (resources+patches+generators), included via \texttt{components:} in overlays. \texttt{resources} = full manifests/bases; \texttt{bases:} key itself is deprecated.
\begin{lstlisting}[language=yaml]
# components/mysql/kustomization.yaml
apiVersion: kustomize.config.k8s.io/v1alpha1
kind: Component
resources: [deployment.yaml, service.yaml]
\end{lstlisting}

\subsection{Best Practices \& Pitfalls}
Keep common values (namespace, labels) in base. Organize by kind (\texttt{services.yaml}, \texttt{hpa.yaml}). Standard dirs: \texttt{base/}, \texttt{overlays/}, \texttt{components/}, \texttt{patches/}.\\
\textbf{Pitfalls:} nested overlays (flatten!); not pinning image tags (use \texttt{images:} with explicit \texttt{newTag}); drift between cluster + kustomization (no release state tracked, unlike Helm); skipping \texttt{kustomize build} validation before apply; using traditional ConfigMaps when rollout-on-change needed (use generator).

\subsection{Helm vs. Kustomize}
\begin{tabular}{ll}
\textbf{Need} & \textbf{Tool} \\
Distribute complex app (DB+App) & Helm \\
Env-specific configs (dev/prod) & Kustomize \\
Avoid new templating language & Kustomize \\
Consume community charts & Helm \\
Loops/conditionals/deps & Helm \\
\end{tabular}
