\renewcommand{\chapcolor}{chTerraform}
\section{Terraform}
\textbf{IaC:} managing/provisioning infra through code instead of manual processes. \textbf{Declarative} (TF, what/desired state, implicit flow, stateful, idempotent) vs \textbf{Imperative} (shell/CLI, how/steps, explicit, stateless, not idempotent). TF talks to Provider plugins (RPC) -> Client SDK (HTTP) -> cloud API. \textbf{Resources} (\texttt{resource}) = managed objects. \textbf{Data sources} (\texttt{data}) = read-only lookups. HCL sample:
\begin{lstlisting}[language=hcl]
terraform {
  required_version = ">= 1.0.0"
  required_providers { aws = { source = "hashicorp/aws", version = "~> 5.0" } }
}
provider "aws" { region = "us-east-1" }
variable "instance_type" { default = "t2.micro" }
resource "aws_instance" "web" {
  ami           = data.aws_ami.ubuntu.id
  instance_type = var.instance_type
}
output "public_ip" { value = aws_instance.web.public_ip }
\end{lstlisting}
Workflow: write \texttt{.tf} -> \texttt{init} (backend + download providers/modules) -> \texttt{plan} -> \texttt{apply} -> \texttt{destroy}. Other: \texttt{fmt}, \texttt{validate}, \texttt{console} (REPL), \texttt{state list/show <addr>}. Reference attrs as \texttt{<type>.<name>.<attr>}.
\subsection{OpenTofu}
Fork of Terraform v1.5.7 under \textbf{Linux Foundation}, MPL-2.0 (TF switched to BUSL). Drop-in replacement, adds \textbf{state encryption}. Avoids single-vendor lock-in.
\subsection{HCL Features}
\textbf{count} (numeric loop, \texttt{count.index}) and \textbf{for\_each} (map/set loop, \texttt{each.key/value}) for multiple instances. \textbf{Data source} example:
\begin{lstlisting}[language=hcl]
data "aws_ami" "ubuntu" {
  most_recent = true
  filter { name = "name"   values = ["ubuntu/images/hvm-ssd/ubuntu-*"] }
  filter { name = "virtualization-type" values = ["hvm"] }
  owners = ["099720109477"] # Canonical
}
\end{lstlisting}
\subsection{Variables}
Precedence: CLI \texttt{-var}/\texttt{-var-file} > env \texttt{TF\_VAR\_<name>} > \texttt{*.tfvars} > default. No default \& not supplied -> CLI prompts operator. \texttt{TF\_VAR\_} prefix ideal for secrets (\texttt{TF\_VAR\_access\_key=...}). \texttt{depends\_on} forces explicit ordering; implicit deps come from attribute references.
\subsection{State}
\texttt{terraform.tfstate} (JSON) maps HCL to real resource IDs (e.g. ARN). \textbf{3-way diff} on apply: \textbf{NEW} (HCL) vs \textbf{PREVIOUS} (.tfstate) vs \textbf{EXISTING} (cloud refresh) -> decides actions, detects drift. \textbf{Remote backend} (S3) shares state in teams; \textbf{state locking} via \texttt{use\_lockfile} (S3-native, replaces DynamoDB):
\begin{lstlisting}[language=hcl]
terraform {
  backend "s3" {
    bucket       = "my-tfstate"
    key          = "prod/terraform.tfstate"
    region       = "us-east-1"
    use_lockfile = true
  }
}
\end{lstlisting}
\subsection{File \& Folder Layout}
\textbf{Files:} \texttt{terraform.tf} (required\_version/providers), \texttt{backend.tf} (remote state, kept out of VCS), \texttt{providers.tf} (auth/config), \texttt{main.tf} (resources/data), \texttt{variables.tf} (inputs, alphabetical), \texttt{outputs.tf} (returns, alphabetical), \texttt{locals.tf} (derived values), \texttt{override.tf} (loaded last, sparingly). \\
\textbf{One folder per env} (\texttt{dev/}, \texttt{prod/}) -> separate state files so a corruption hits only one env; shared \texttt{modules/} alongside.
\subsection{Modules}
Reusable folder of \texttt{.tf} files (root config is itself a module):
\begin{lstlisting}[language=hcl]
module "vpc" { source = "./modules/vpc"  cidr = "10.0.0.0/16" }
# reference output: module.vpc.vpc_id
\end{lstlisting}
\subsection{Patterns \& Recovery}
\texttt{import} adopts existing infra; \texttt{terraform apply -replace=<addr>} (or legacy \texttt{taint}) forces a destroy+recreate of a single resource. Region-aware AMI via map:
\begin{lstlisting}[language=hcl]
import { to = aws_instance.web   id = "i-0abc123" }   # then: apply

variable "AMIS" { type = map(string)  default = { us-east-1 = "ami-05b1..." } }
ami = var.AMIS[var.region]
\end{lstlisting}
